---
title: "VI with Stan"
editor: 
  mode: source
---

{{< include ../_macros.qmd >}}

## Outline

### Topics

-   Calling VI with cmdstanr
-   Main options
-   Interpreting the output
-   Alternatives to Stan

### Rationale

Now that you know how to write Stan models, we see here that you can run not only 
HMC but also VI. 


## Example

We will use the [model for detecting random guessing](../w10_modelling/topic07_mixtures.qmd) 
and compare the output of variational inference and HMC. 


## Calling VI with cmdstanr

The commands are very similar to how we call HMC. 

First create the model exactly as before:

```{r message=FALSE, warning=FALSE, results=FALSE}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(bayesplot))

data = read.csv(url("https://github.com/UBC-Stat-ML/web447/raw/main/data/ScoreData.csv"))
mod = cmdstan_model("../w10_modelling/students_guessing.stan")
```

Then to run VI: 

```{r}
fit.variational = mod$variational(
  seed = 1,
  refresh = 500,
  output_dir = "stan_out",
  algorithm = "meanfield", # <1>
  output_samples = 1000, # <2>
  data = list(
            n_students = length(data$Score),
            scores = data$Score
          ) 
)
```
1. This specifies the variational family $\family$. Use `meanfield` for a normal with diagonal covariance, `fullrank` for any normal. 
2. How many samples from $q^*$ to output.

**Note:** calling `mod$variational` does two things:

1. First, find $q^* \in \family$ that is closest to $\pi$. 
2. Sample from $q^*$ `output_samples` times (this provides independent and identical samples). 


Compare this to how we ran HMC:

```{r}
fit.hmc = mod$sample(
  seed = 1,
  refresh = 500,
  output_dir = "stan_out",
  chains = 1,
  data = list(
            n_students = length(data$Score),
            scores = data$Score
          ) 
)
```



## Comparing the output of VI and HMC

Recall that in [the mixture page we defined](../w10_modelling/topic07_mixtures.qmd) random variables $X_i = \ex[G_i | A_i, Y]$, the 
"guessing probabilities", with support on $[0, 1]$. 

Let us look at the posteriors of $X_i$ for HMC vs VI:

```{r}
#| code-fold: true
#| warning: false

library(bayesplot)
library(ggplot2)
library(dplyr)
library(latex2exp)

vars = fit.hmc$metadata()$model_params
sel  = grep("^guessing_probabilities.*", vars, value = TRUE)

data_hmc = mcmc_intervals_data(fit.hmc$draws(variables = sel)) %>% 
  mutate(method = "HMC")

data_vi = mcmc_intervals_data(fit.variational$draws(variables = sel)) %>% 
  mutate(method = "Variational")

bind_rows(data_hmc, data_vi) %>%
  ggplot(aes(x = m, y = parameter, color = method)) +
  geom_linerange(aes(xmin = ll, xmax = hh), 
                 position = position_dodge(width = 0.5)) +
  geom_linerange(aes(xmin = l, xmax = h), 
                 position = position_dodge(width = 0.5), size = 1.5) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_x_continuous(limits = c(0, 1)) +
  theme_minimal() +
  labs(
    x = TeX("$X_i$"),
    y = NULL,
    color = "Method"
  )
```

**Question:** If the variational family is normal, how come the support of $X_i$  seems to be contained in the interval $[0, 1]$? 

{{< include ../../clickers/w13/_c02.qmd >}}