---
title: "Grouped data"
editor: 
  mode: source
---

{{< include ../_macros.qmd >}}

## Outline 

### Topics

- Using the package `tidybayes` to feed complex data into Stan.

### Rationale

Data organized in various groups is a frequent feature of Bayesian models, 
in particular in [hierarchical models](../w06_hierarchical/topic03_hierarchy.qmd). 

Getting grouped data into stan is tedious and error-prone.
The package `tidybayes` automates much of this. 


## Pre-reading

If you have never heard about "tidy data", while it is not strictly essential for this course, 
it is a good investment to skim this [tutorial on tidy data](https://r4ds.had.co.nz/tidy-data.html). 

## Example

First, install the packages `magrittr` and `tidybayes` (and `ggplot2` if you have not done so already), 
then import them:

```{r}
suppressPackageStartupMessages(require(cmdstanr))
suppressPackageStartupMessages(require(magrittr))
suppressPackageStartupMessages(require(tidybayes))
suppressPackageStartupMessages(require(ggplot2))
```


### Data prep

We load data modelling covid vaccines efficacy:

```{r}
data = read.csv(url("https://raw.githubusercontent.com/UBC-Stat-ML/web447/main/exercises/ex06_assets/vaccines_full.csv"))
data$is_vaccinated = ifelse(data$arms == "vaccinated", 1, 0)
rmarkdown::paged_table(data)
```

The magic conversion of "tidy data" (data in a format like the above) into a format
that can be consumed by Stan is done using `compose_data`:

```{r}
stan_converted = compose_data(data)

stan_converted
```


### Stan model 

Here we consider a simple, non-hierarchical model but fitting all the data 
at once. As an exercise, think about how you can modify this model to be hierarchical. 

```stan {shortcodes=true, filename="vaccines.stan"}
{{< include vaccines.stan >}}
```

### Fitting

The fit object returned by `sample()` does not know about the string labels attached to
each trial integer index. We use `fit %<>% recover_types(data)` to add back that information:

```{r message=FALSE, warning=FALSE, results=FALSE}
mod = cmdstan_model("vaccines.stan")
fit = mod$sample(
  seed = 1,
  data = stan_converted, 
  output_dir = "stan_out",
  refresh = 0,
  iter_sampling = 10000                  
)
fit %<>% recover_types(data)
```



### Getting draws

The output of Stan is not tidy either. 

```{r}
fit
```


Use `spread_draws` to put the draws into a tidy format:

```{r}
fit %>% spread_draws(efficiencies[trials], prevalences[trials]) %>% head(10)
```

### Plotting

Now that we have draws in tidy format, instead of using specialized MCMC plotting 
libraries we can just use ggplot:

```{r}
fit %>%
  spread_draws(efficiencies[trials]) %>%
  ggplot(aes(x = efficiencies, y = trials)) +
  stat_halfeye() + 
  theme_minimal()

```

This makes it easier to customize plots.


### Summaries

`tidybayes` also offers convenient ways to compute summaries such as 
[High Density Intervals (HDI)](../w05_properties/topic04_decision_to_set.qmd)

```{r}
fit %>%
  spread_draws(efficiencies[trials]) %>%
  median_hdi(efficiencies)
```

## More information

See the [tidybayes documentation](https://mjskay.github.io/tidybayes/).