---
title: "Stan: going further"
editor: 
  mode: source
---

{{< include ../_macros.qmd >}}

## Outline

### Topics

-   More complex types in Stan (vectors, constraints) 
-   Loops and vectorization
-   Feeding complex data into Stan 

### Rationale

We review here the constructs needed to write more complex models such as 
the exercises coming in the 
next few weeks, and the second quiz. 


## Example 

The data below consists, for each day in December 2021,
of the **percentage** of sequenced COVID-19 samples that are of the
Omicron variant.


```{r}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))

# data from: https://data.chhs.ca.gov/dataset/covid-19-variant-data
df = read.csv("../data/covid19_variants.csv")
df$date = as.Date(df$date,format="%Y-%m-%d")

df = df %>% filter(date > "2021-12-01" & date < "2021-12-31") %>% filter(variant_name == "Omicron")

df %>% ggplot(aes(x = date, y = percentage)) + geom_point() + ylim(0, 100) + theme_minimal()
```


## Model (mathematical notation)

One possible model for this data: for $i \in \{1, 2, \dots, N\}$, $N$ being the number of days in December ($N=31$),[^1]

[^1]: In the model below, notice we normalize the day by $N$ to put it in the interval $[0, 1]$. This is not strictly necessary, but often,
  bringing covariates into a nice range can make sampling faster when it reduces the 
  [condition number of the sampling problem](https://mc-stan.org/docs/stan-users-guide/efficiency-tuning.html#unit-scales-without-correlation).

\begin{align*}
\theta_0 &\sim \distExp(1) \\
\theta_1 &\sim \distNorm(0, \sigma = 1000) \\
\theta_2 &\sim \distNorm(0, \sigma = 1000) \\
\mu_i &= \text{logistic}(\theta_1 (i/N) + \theta_2) \\
y_i &\sim \distBeta(\mu_i, \theta_0) \\
\end{align*}


## Model (in Stan)

Let us translate the mathematical notation into Stan:


```stan {shortcodes=true, filename="omicron_first_attempt.stan"}
{{< include omicron_first_attempt.stan >}}
```
1. Here, `beta_proportion` corresponds to the [mean-concentration parameterization](https://mc-stan.org/docs/functions-reference/continuous_distributions_on_0_1.html#beta-proportion-distribution) we discussed [in the lecture on hierarchical models](../w06_hierarchical/topic03_hierarchy.qmd).

**Question:** how do you think Stan will use the parameters' type constraints? (e.g., `<lower=0>` in `real<lower=0> concentration`)?

{{< include ../../clickers/w08/_c01.qmd >}}



## Improving our model with transformed parameters

- The code in the last section will not print out $\mu_i$ 
- The code below shows how to fix this
    - $\mu_i$ is handled differently than $\theta_k$, $y_i$...
    - ... because $\mu_i$ is defined with $=$ instead of $\sim$


```stan {shortcodes=true, filename="omicron.stan"}
{{< include omicron.stan >}}
```
1. Whenever in the mathematical notation a latent random variable is defined using an **equality** ("$=$"), and you want to have it as part of 
the output, write it in the `transformed parameters` block.
2. Whenever in the mathematical notation a latent random variable is defined using a **distribution statement** ("$\sim$"), write it in the `model` block. 


## Running MCMC

Running MCMC on the model is done as before:[^2] 

[^2]: note that we make sure no datapoints have value zero or one as under certain parameter values of the beta distribution, this can lead to a zero density point which Stan is not able to handle. 

```{r message=FALSE, warning=FALSE, results=FALSE}
data_without_zeros = pmax(pmin(df$percentage/100,0.999),0.001)

mod = cmdstan_model("omicron.stan")
fit = mod$sample(
  data = list( # <1>
          y = data_without_zeros, 
          N = length(data_without_zeros)
        ), 
  chains = 1, # <2>
  iter_sampling = 10000,
  refresh = 5000,
  output_dir = "stan_out",
)
```
1. Each variable in Stan's `data` block needs to be fed a value when called from R.
2. The `chains` option can be used to simulate several independent MCMC chains, moreover, they 
    can be sampled in parallel using instead `parallel_chains` in `mod$sample(...)`.


## Posterior visualization

We can extract samples as follows:

```{r}
samples = posterior::as_draws_matrix(fit$draws("mu"))
n_samples = nrow(samples)
n_samples
```

and plot the posterior distribution over the beta's mean functions:

```{r}
data = df$percentage
xs = 1:length(data) / length(data)

plot(xs, data/100, 
      xlab = "Fraction of the month of December, 2021", 
      ylab = "Omicron fraction")

for (i in 1:n_samples) {
  lines(xs, samples[i,], col = rgb(red = 0, green = 0, blue = 0, alpha = 0.01))
}
```

 

## Model criticism 

What are potential weakness(es) of the model? 

{{< include ../../clickers/w08/_c02.qmd >}}


