---
title: "Mixtures"
editor: 
  mode: source
---

{{< include ../_macros.qmd >}}


## Outline

### Topics

- Mathematical definition of mixtures
- Implementation in Stan


### Rationale

Mixtures are motivated by cases where you believe the data is composed of 
two or more distinct sub-populations. 

One use of mixtures is to figure out, for each data point, from which 
sub-population it comes from. 

Beyond the sub-population example, mixtures are also used to get more flexible 
distribution families. 


## Example: detection of random guessing

- We look at a dataset from [Albert and Hu, 2020](https://bayesball.github.io/BOOK/case-studies.html#latent-class-modeling). 
- We have scores for 30 students on a test. 
- The test has 20 questions. 
- All questions are TRUE/FALSE with same weight. 
- Every student attempted every question. 

```{r}
suppressPackageStartupMessages(require("rstan"))

data = read.csv(url("https://github.com/UBC-Stat-ML/web447/raw/main/data/ScoreData.csv"))
rmarkdown::paged_table(data)
```

```{r}
plot(data$Score,
      ylim = c(0, 20),
      xlab = "Student index", 
      ylab = "Score (out of 20 questions)")
```


- Notice a cluster of students with scores around 50%,
- and a second cluster with much better scores. 
- Hypothesis: some students answered completely at random.


## Simplified problem

- Suppose first that we have the score for only one student.
- E.g.: first one, who got 9/20. 
- You want a model to put a probability on: "is this student is guessing at random?"
    - Real application: analysis of clicker scores where grade is participation-based. 


### Mathematical model

- Start with the data: $Y$, an integer between 0 and 20. 

**Question:** what is a good choice of likelihood?

{{< include ../../clickers/w10/_c05.qmd >}}

- The parameters of the likelihood depend on whether the student is guessing or not
    - Let $G$ denote an indicator on guessing, i.e.:
        - $G = 1$: guessing,
        - $G = 0$: not guessing, i.e., trying at the best of their ability. 
        
        
**Question:** what is a good choice of prior on $G$?

{{< include ../../clickers/w10/_c06.qmd >}} 


- The likelihood given $G = 1$ (guessing) is easy:
  $$Y \sim \distBinom(20, 0.5).$$
- The likelihood given $G = 0$ (trying at the best of their ability) is slightly trickier:
    - what is "the best of their ability"?
    - **Bayesian recipe:** treat it as an unknown ability parameters $A$
    
**Full model:**

\begin{align*}
A &\sim \distUnif(0, 1) \\
G &\sim \distBern(1/3) \\
Y &\sim \distBinom(20, 0.5 G + A(1-G)).
\end{align*}


### Marginalization

- Stan does not support discrete latent random variables like $G$
- Therefore we use [Rao-Blackwellization](topic06_rao.qmd):
    - Joint distribution before Rao-Blackwellization:
      $$\gamma(a, g, y) = p(a) p(g) p(y|a, g).$$
    - Joint distribution after Rao-Blackwellization:
      \begin{align*}
      \gamma(a, y) &= \sum_{g=0}^1 \gamma(a, g, y) \\
      &= \sum_{g=0}^1 p(a) p(g) p(y|a, g) \\
      &= p(a) \sum_{g=0}^1 p(g) p(y|a, g) \\
      &= \underbrace{p(a)}_\text{(a)} \underbrace{\left( \underbrace{p(g=1) p(y|a, g=1)}_\text{(b)} + \underbrace{p(g=0) p(y|a, g=0)}_\text{(c)} \right)}_\text{(d)}.
      \end{align*}
      
**Question:** match-up (a)-(d) with the labelled statements in the Stan code below. 


{{< include ../../clickers/w10/_c07.qmd >}}


```{stan output.var = "students_guessing_simplified"}
data {
  int<lower=0, upper=20> score;
}

parameters {
  real<lower=0, upper=1> ability;
}

transformed parameters {
  real complete_likelihood_guessing      = 1.0/3 * exp(binomial_lpmf(score | 20, 0.5)); # <1>
  real complete_likelihood_non_guessing  = 2.0/3 * exp(binomial_lpmf(score | 20, ability)); # <2>
}

model {
  ability ~ uniform(0, 1); # <3>
  target += log(complete_likelihood_guessing + complete_likelihood_non_guessing); # <4>
}

generated quantities {
  real guessing_probability = # <5>
    complete_likelihood_guessing / (complete_likelihood_guessing + complete_likelihood_non_guessing);
}
```
1. Stan statement 1 (used in clicker question)
2. Stan statement 2
3. Stan statement 3
4. Stan statement 4
5. Stan statement 5


### Re-instatiation

- We care about the posterior mean of $G$, $\ex[G | Y = y]$...
- ...but we have marginalized $G$ 🙁
- To compute the posterior mean, we first use the law of total expectation:
  $$\ex[G | Y] = \ex[\red{\ex[G | A, Y]} | Y],$$
- Let us look at the part in red, and see how we can compute it at every MCMC iteration based on the current value of $A^\parm$...
    - Using the fact the mean of a Bernoulli is the probability that it takes value 1:
      $$\ex[G | A, Y] = \pr(G = 1 | A, Y).$$
    - By Bayes rule:
      $$\pr(G = 1 | A, Y) = \frac{\pr(G = 1 | A, Y)}{\pr(G = 0 | A, Y) + \pr(G = 1 | A, Y)}.$$
    - This is what gets computed in statement ⑤ in the above Stan code. 
- Now, recall from the [LLN for Markov chains](../w08_mcmc1/topic05_mcmc_consistency.qmd) gives us condition so that for a test function $h$, we have that the Monte Carlo samples, $A^{(1)}, A^{(2)}, \dots, A^{(M)}$ satisfy,
$$\frac{1}{M} \sum_{m=1}^M h(A^\parm) \to \ex[h(A) | Y],$$
with probability one.
- Therefore from taking $h(a) = \pr(G = 1 | A = a, Y)$, we obtain,
$$\frac{1}{M} \sum_{m=1}^M \pr(G = 1 | A^\parm, Y = y) \to \ex[G | Y],$$
with probability one.


### Testing the model

**Probably guessing:** feeding a score of **9** out of 20, we get:


```{r dependson=knitr::dep_prev()}
fit = sampling(
  students_guessing_simplified,
  seed = 1,
  refresh = 0,
  data = list(score = 9)       
)

mean(extract(fit)$guessing_probability)
```

**Probably not guessing:** feeding a score of **19** out of 20, we get:

```{r dependson=knitr::dep_prev()}
fit = sampling(
  students_guessing_simplified,
  seed = 1,
  refresh = 0,
  data = list(score = 19)      
)

mean(extract(fit)$guessing_probability)
```


**What is the key limitation of this approach?**

::: {.callout-caution collapse="true"} 
## Click for answer

The student could have a low score because the questions were too difficult! 

With some assumptions (unimodality of the students abilities) we can avoid this limitation.
:::



## Full model: getting rid of key limitation

- **Idea:** use a [hierarchical model](../w06_hierarchical/topic03_hierarchy.qmd)! 
- Random variables that are student-specific:
    - $Y_i \sim \distBinom(20, 0.5 G_i + A_i (1 - G_i)).$
    - $A_i \sim \distBeta(\mu, S)$
    - $G_i \sim \distBern(F)$
- Random variables that are global:
    - Population fraction that are guessing: $F \sim \distUnif(0, 1)$,
    - Population parameters for the abilities:
        - $\mu \sim \distUnif(0, 1)$ 
        - $S \sim \distExp(1/100)$. 
- Stan implementation:

```{stan output.var = "students_guessing"}
data {
  int n_students; 
  array[n_students] int<lower=0, upper=20> scores; 
}

parameters {
  real<lower=0, upper=1> fraction_guessing;
  real<lower=0, upper=1> non_guessing_population_mean;
  real<lower=0> non_guessing_population_spread;
  vector<lower=0, upper=1>[n_students] abilities;
}

transformed parameters {
  vector[n_students] complete_loglikelihood_guessing; 
  vector[n_students] complete_loglikelihood_non_guessing; 
  for (i in 1:n_students) {
    complete_loglikelihood_guessing[i]     = log(fraction_guessing)    + binomial_lpmf(scores[i] | 20, 0.5); 
    complete_loglikelihood_non_guessing[i] = log1p(-fraction_guessing) + binomial_lpmf(scores[i] | 20, abilities[i]);
  }
}

model {
  fraction_guessing ~ uniform(0, 1);
  non_guessing_population_mean ~ uniform(0, 1);
  non_guessing_population_spread ~ exponential(1.0/100);
  for (i in 1:n_students) {
    abilities[i] ~ beta_proportion(non_guessing_population_mean, non_guessing_population_spread);
    target += log_sum_exp(complete_loglikelihood_guessing[i], complete_loglikelihood_non_guessing[i]);
  }
}

generated quantities {
  vector[n_students] guessing_probabilities = inv_logit(complete_loglikelihood_guessing - complete_loglikelihood_non_guessing);
}
```

```{r  message=FALSE, warning=FALSE, dependson=knitr::dep_prev()}
fit = sampling(
  students_guessing,
  seed = 1,
  refresh = 0,
  data = list(
            n_students = length(data$Score),
            scores = data$Score
          )                  
)
```

**Credible intervals and posterior medians:** 90% (thick lines) and 50% (thin lines)
 
```{r message=FALSE, warning=FALSE}

mcmc_intervals(fit, regex_pars = c("guessing_probabilities.*")) + 
  theme_minimal() + 
  scale_x_continuous(limits = c(0, 1)) 
```