---
title: "Mixtures"
editor: 
  mode: source
---

{{< include ../_macros.qmd >}}


## Outline

### Topics

- Mathematical definition of mixtures
- Implementation in Stan


### Rationale

Mixtures are motivated by cases where you believe the data is composed of 
two or more distinct sub-populations. 

One use of mixtures is to figure out, for each data point, from which 
sub-population it comes from. 

Beyond the sub-population example, mixtures are also used to get more flexible 
distribution families. 


## Example: detection of random guessing

- We look at a dataset from [Albert and Hu, 2020](https://bayesball.github.io/BOOK/case-studies.html#latent-class-modeling). 
- We have scores for 30 students on a test. 
- The test has 20 questions. 
- All questions are TRUE/FALSE with the same weight. 
- Every student attempted every question. 

```{r}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(bayesplot))

data = read.csv(url("https://github.com/UBC-Stat-ML/web447/raw/main/data/ScoreData.csv"))
rmarkdown::paged_table(data)
```

```{r}
plot(data$Score,
      ylim = c(0, 20),
      xlab = "Student index", 
      ylab = "Score (out of 20 questions)")
```


- Notice a cluster of students with scores around 50%,
- and a second cluster with much better scores. 
- Hypothesis: some students answered completely at random.


## Simplified problem

- Suppose first that we have the score for only one student.
- E.g.: first one, who got 9/20. 
- You want a model to put a probability on: "is this student guessing at random?"
    - Real application: analysis of clicker scores where the grade is participation-based. 


### Mathematical model

- Start with the data: $Y$, an integer between 0 and 20. 

**Question:** what is a good choice of likelihood?

{{< include ../../clickers/w10/_c05.qmd >}}

- The parameters of the likelihood depend on whether the student is guessing or not
    - Let $G$ denote an indicator on guessing, i.e.:
        - $G = 1$: guessing,
        - $G = 0$: not guessing, i.e., trying at the best of their ability. 
        
        
**Question:** what is a good choice of prior on $G$?

{{< include ../../clickers/w10/_c06.qmd >}} 


- The likelihood given $G = 1$ (guessing) is easy:
  $$Y \sim \distBinom(20, 0.5).$$
- The likelihood given $G = 0$ (trying at the best of their ability) is slightly trickier:
    - what is "the best of their ability"?
    - **Bayesian recipe:** treat it as an unknown ability parameter $A$
    
**Full model:**

\begin{align*}
A &\sim \distUnif(0, 1) \\
G &\sim \distBern(1/3) \\
Y &\sim \distBinom(20, 0.5 G + A(1-G)).
\end{align*}


### Marginalization

- Stan does not support discrete latent random variables like $G$.
- Therefore we use [Rao-Blackwellization](topic06_rao.qmd):
    - Joint distribution before Rao-Blackwellization:
      $$\gamma(a, g, y) = p(a) p(g) p(y|a, g).$$
    - Joint distribution after Rao-Blackwellization:
      \begin{align*}
      \gamma(a, y) &= \sum_{g=0}^1 \gamma(a, g, y) \\
      &= \sum_{g=0}^1 p(a) p(g) p(y|a, g) \\
      &= p(a) \sum_{g=0}^1 p(g) p(y|a, g) \\
      &= \underbrace{p(a)}_\text{(a)} \underbrace{\left( \underbrace{p(g=1) p(y|a, g=1)}_\text{(b)} + \underbrace{p(g=0) p(y|a, g=0)}_\text{(c)} \right)}_\text{(d)}.
      \end{align*}
      
**Question:** match-up (a)-(d) with the labelled statements in the Stan code below. 


{{< include ../../clickers/w10/_c07.qmd >}}


```stan {shortcodes=true, filename="students_guessing_simplified.stan"}
{{< include students_guessing_simplified.stan >}}
```
1. Stan statement 1 (used in clicker question)
2. Stan statement 2
3. Stan statement 3
4. Stan statement 4
5. Stan statement 5


### Re-instantiation

- We care about the posterior mean of $G$, $\ex[G | Y = y]$...
- ...but we have marginalized $G$ 🙁
- To compute the posterior mean, we first use the law of total expectation:
  $$\ex[G | Y] = \ex[\red{\ex[G | A, Y]} | Y],$$
- Let us look at the part in red, and see how we can compute it at every MCMC iteration based on the current value of $A^\parm$...
    - Using the fact that the mean of a Bernoulli is the probability that it takes value 1:
      $$\ex[G | A, Y] = \pr(G = 1 | A, Y).$$
    - By Bayes' rule (noting that a conditional probability like $\pr(\cdot | A)$ [is a probability](../w01_discrete_inference/topic07_conditional.qmd#a-conditional-probability-is-a-probability)):
      $$\pr(G = 1 | A, Y = y) = \frac{\pr(G = 1, Y = y | A)}{\pr(G = 0, Y = y | A) + \pr(G = 1, Y = y | A)}.$$
    - This is what gets computed in statement ⑤ in the above Stan code. 
- Now, recall from the [LLN for Markov chains](../w08_mcmc1/topic05_mcmc_consistency.qmd) gives us conditions so that for a test function $h$, we have that the Monte Carlo samples, $A^{(1)}, A^{(2)}, \dots, A^{(M)}$ satisfy,
$$\frac{1}{M} \sum_{m=1}^M h(A^\parm) \to \ex[h(A) | Y = y],$$
with probability one.
- Therefore from taking $h(A^\parm) = \pr(G = 1 | A^\parm, Y = y)$, we obtain,
$$\frac{1}{M} \sum_{m=1}^M \pr(G = 1 | A^\parm, Y = y) \to \ex[G | Y = y],$$
with probability one.


### Testing the model

**Probably guessing:** feeding a score of **9** out of 20, we get:

<!-- we need to show the output of this one! -->
```{r warning=FALSE}
mod = cmdstan_model("students_guessing_simplified.stan")
fit = mod$sample(
  seed = 1,
  refresh = 0,
  output_dir = "stan_out",
  chains = 1,
  data = list(score = 9)       
)

guessing_probability_samples = as.vector(fit$draws("guessing_probability"))
mean(guessing_probability_samples)
```


**Probably not guessing:** feeding a score of **19** out of 20, we get:

<!-- we need to show the output of this one! -->
```{r warning=FALSE}
fit = mod$sample(
  seed = 1,
  refresh = 0,
  output_dir = "stan_out",
  chains = 1,
  data = list(score = 19)       
)

guessing_probability_samples = as.vector(fit$draws("guessing_probability"))
mean(guessing_probability_samples)
```


**What is the key limitation of this approach?**

::: {.callout-caution collapse="true"} 
## Click for answer

The student could have a low score because the questions were too difficult for them! 

With some assumptions (unimodality of the students abilities) we can avoid that limitation.
:::



## Full model: getting rid of the key limitation

- **Idea:** use a [hierarchical model](../w06_hierarchical/topic03_hierarchy.qmd)! 
- Random variables that are student-specific:
    - $Y_i \sim \distBinom(20, 0.5 G_i + A_i (1 - G_i)).$
    - $A_i \sim \distBeta(\mu, S)$
    - $G_i \sim \distBern(F)$
- Random variables that are global:
    - Population fraction of students who are guessing: $F \sim \distUnif(0, 1)$,
    - Population parameters for the abilities:
        - $\mu \sim \distUnif(0, 1)$ 
        - $S \sim \distExp(1/100)$. 
- Stan implementation:


```stan {filename="students_guessing.stan"}
{{< include students_guessing.stan >}}
```

```{r message=FALSE, warning=FALSE, results=FALSE}
mod = cmdstan_model("students_guessing.stan")
fit = mod$sample(
  seed = 1,
  refresh = 0,
  output_dir = "stan_out",
  chains = 1,
  data = list(
            n_students = length(data$Score),
            scores = data$Score
          ) 
)
```



**Credible intervals and posterior medians:** 90% (thick lines) and 50% (thin lines), for the random variables $\ex[G_i | A_i, Y]$
 
```{r message=FALSE, warning=FALSE}

vars <- fit$metadata()$model_params
sel  <- grep("^guessing_probabilities.*", vars, value = TRUE)
mcmc_intervals(fit$draws(variables = sel)) + 
  theme_minimal() + 
  scale_x_continuous(limits = c(0, 1)) 
```

**Predictive score distribution:** for the non guessing sub-population

```{r message=FALSE, warning=FALSE}

mcmc_areas_ridges(fit$draws(c("predictive_score_non_guessing"))) + 
  theme_minimal() + 
  scale_x_continuous(limits = c(0, 20)) 
```