---
title: "Exercise 6: Hierarchical models"
editor: 
  mode: source
---
  
{{< include ../_macros.qmd >}}


## Q.1: efficacy of vaccines

In this exercise we will model the effectiveness of COVID vaccines using clinical
trials data.
Each trial consists of two *arms*: *vaccinated* (i.e., *treated*) and *control*
(i.e., *not* treated). For a typical trial we know

a. The total number of patients in each arm: $t_\text{v}$ (vaccinated), $t_\text{c}$ (control).
b. The number of patients that got infected with the SARS-CoV-2 virus in each
arm: $n_\text{v}$, $n_\text{c}$.

We model $n_\text{v}$ and $n_\text{c}$ as Binomial random variables. The unknown parameter for
these distributions will depend on two numbers in $[0,1]$

a. Incidence $p$: the probability that a patient in the trial
will become infected without being treated with the vaccine.
a. Effectiveness $e$: the decrease in incidence that the vaccine provides.

We put a beta prior on both parameters. Since the parameters of the
$\distBeta(\alpha,\beta)$ distribution are not very interpretable, we will
work with a re-parametrized version $\distNamed{BetaMP}(\mu,\lambda)$ where 
$\mu\in[0,1]$ is the *mean* and $\lambda>0$ is a *precision* parameter. The
bijection is given by
$$
\mu = \frac{\alpha}{\alpha+\beta} \quad \lambda=\alpha+\beta \qquad\iff\qquad
\alpha = \mu \lambda \quad \beta = (1-\mu)\lambda
$$

Finally, we put priors on the parameters of these beta distributions to complete
the model as follows
$$
\begin{aligned}
\mu_\text{e} &\sim \distUnif(0,1) && (\text{diffuse prior})\\
\lambda_\text{e} &\sim \distExp(0.01) && (\text{diffuse prior})\\
\mu_\text{p} &\sim \distNamed{BetaMP}(0.1, 15) && (\text{incidence is usually low})\\
\lambda_\text{p} &\sim \distExp(0.01) && (\text{diffuse prior})\\
e|\mu_\text{e},\lambda_\text{e} &\sim \distNamed{BetaMP}(\mu_\text{e}, \lambda_\text{e}) \\
p|\mu_\text{p},\lambda_\text{p} &\sim \distNamed{BetaMP}(\mu_\text{p}, \lambda_\text{p}) \\
n_\text{c}|p &\sim \distBinom(t_\text{c}, p) \\
n_\text{v}|e,p &\sim \distBinom(t_\text{v}, p(1-e))
\end{aligned}
$$ {#eq-model-one-vac}



```{r}
#| echo: false
dta = read.csv("ex06_assets/vaccines.csv")
```

Now, consider the following 
data---available [here](ex06_assets/vaccines.csv)---arising from clinical trials 
of two popular COVID vaccines
```{r}
#| echo: false
rmarkdown::paged_table(dta)
```

1. Expand the model in @eq-model-one-vac into a hierarchical model that covers 
both vaccines. The parameters $(\mu_\text{e},\lambda_\text{e},\mu_\text{p},\lambda_\text{p})$ must be 
shared across vaccines. In contrast, each vaccine must have its own $(e,p)$ 
pair.
2. Implement the model from part 1 in [simPPLe](../w04_glms/topic01_simPPLe.qmd).
Your PPL function should return the indicator that Moderna is more effective than
Pfizer. Use this code to define $\distNamed{BetaMP}(\mu, \lambda)$
```r
BetaMP = function(mean, precision){
  Beta(mean*precision, (1-mean)*precision)
}
```
3. Download the [data](ex06_assets/vaccines.csv) and load it using the following
(no points for this part)
```r
vaccines = read.csv("vaccines.csv")
vaccines$groupSizes = as.double(vaccines$groupSizes) # needed due to bug in Binom code
```
4. Run your model in [simPPLe](../w04_glms/topic01_simPPLe.qmd) using `posterior`
with 10,000 iterations, and report the estimated posterior probability that 
Moderna is more effective than Pfizer. Is the SNIS approximation reliable?