---
title: "Kernel alternations"
editor: 
  mode: source
---

{{< include ../_macros.qmd >}}

## Outline

### Topics

- Alternations of MCMC kernels
- Code implementation
- Alternations preserve invariance

### Rationale

*Alternation of kernels* is another method to combine kernels. 
It is mathematically slightly more complicated than mixtures, but can yield 
faster mixing (in some situations, [much faster](https://rss.onlinelibrary.wiley.com/doi/10.1111/rssb.12464)).

Here we just give an introduction to kernel alternation.
We will talk about the reason why they can yield much faster performance when 
we talk about *non-reversibility.*


## Alternation of MCMC kernels

### Example 

We use the same example as the [one we used for MCMC mixtures](topic06_mix.qmd).

```{r}
#| eval: true
#| code-fold: true
#| file: ../exercises/ex07_scaffold.R
```

### Intuition

**Setup:** Denote the 2 kernels we want to use by $K_1$ and $K_2$. 

**Idea:** at each MCMC iteration, let $x$ denote the current state, 

1. Sample from $K_1$ from the current state.
2. Sample from $K_2$ from the state output by $K_1$. 


### Coding up kernel alternation

Here is an example how to implement the alternation.
It is based on the same normal proposal MH kernels, $K_1$ and $K_2$, with 
standard deviations 1 and 2 respectively.

```{r}
#| eval: true
#| code-fold: true
#| file: ../blocks/normal_mh_kernel.R
```

```{r}

# simple Metropolis-Hastings algorithm (normal proposal)
mcmc_mixture = function(gamma, initial_point, n_iters) {
  samples = numeric(n_iters) 
  dim = length(initial_point)
  current_point = initial_point
  for (i in 1:n_iters) {

    intermediate_point = kernel(gamma, current_point, 1)
    current_point = kernel(gamma, intermediate_point, 2)
    
    samples[i] = current_point
  }
  return(samples)
}

samples = mcmc_mixture(gamma_beta_binomial, 0.5,  1000)
```


### Mathematical notation for kernel mixtures

**Question:** write a mathematical expression for the alternation $\Kalt$ in terms of $K_1$ and $K_2$. 

{{< include ../../clickers/w12/_c08.qmd >}} 


## Invariance result

### Alternation preserves invariance

**Proposition:** if $K_i$ are $\pi$-invariant, then their alternation is $\pi$-invariant. 

**Proof:**

\begin{align*}
\sum_{x} \pi(x) \Kalt(x' | x) &= \sum_{x} \pi(x) \sum_{\check x} K_1(\check x | x) K_2(x' | \check x) \\
&=  \sum_{\check x} K_2(x' | \check x) \sum_{x} \pi(x)  K_1(\check x | x) \\
&= \sum_{\check x} K_2(x' | \check x) \pi(\check x) \expl{invariance of $K_1$} \\
&= \pi(x') \expl{invariance of $K_2$}.
\end{align*}
